---
title: "Class 11: Alphafold"
author: "Qiman A17912409"
format: pdf
---

## Custom analysis of resulting models

For tidiness we can move our AlphaFold results directory into our RStudio project directory. 

```{r}
results_dir <- "hivprdimer_23119.result"
```

```{r}
#File names for all PDB models
pdb_files <- list.files(
  path = results_dir,
  pattern = "\\.pdb$",
  full.names = TRUE,
  recursive = TRUE
)

# Print our PDB file names
basename(pdb_files)
```

```{r}
library(bio3d)

# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```

```{r}
pdbs
```

RMSD is a standard measure of structural distance between coordinate sets. We can use the `rmsd()` function to calculate the RMSD between all pairs models.

```{r}
rd <- rmsd(pdbs, fit=T)
range(rd)
```

Draw a heatmap of these RMSD matrix values:

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

Now lets plot the pLDDT values across all models. Recall that this information is in the B-factor column of each model and that this is stored in our aligned `pdbs` object as `pdbs$b` with a row per structure/model.
```{r}
# Read a reference PDB structure
pdb <- read.pdb("1hsg")
```


You could optionally obtain secondary structure from a call to `stride()` or `dssp()` on any of the model structures.

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```
We can improve the superposition/fitting of our models by finding the most consistent “rigid core” common across all the models. For this we will use the `core.find()` function:

```{r}
core <- core.find(pdbs)

```

```{r}
core.inds <- print(core, vol=0.5)
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

Now we can examine the RMSF between positions of the structure. RMSF is an often used measure of conformational variance along the structure:

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```


Independent of the 3D structure, AlphaFold produces an output called **Predicted Aligned Error (PAE)**. This is detailed in the JSON format result files, one for each model structure.

Below we read these files and see that AlphaFold produces a useful inter-domain prediction for model 1 (and 2) but not for model 5 (or indeed models 3, 4, and 5):

```{r}
library(jsonlite)

results_dir <- "hivprdimer_23119.result"

# List the model score JSON files only
pae_files <- list.files(
  path = results_dir,
  pattern = "scores_rank_.*model.*\\.json$",
  full.names = TRUE,
  recursive = TRUE
)

basename(pae_files)
length(pae_files)
```

```{r}
pae1 <- read_json(pae_files[1], simplifyVector = TRUE)
pae5 <- read_json(pae_files[5], simplifyVector = TRUE)

attributes(pae1)
```

```{r}
# Per-residue pLDDT scores 
#  same as B-factor of PDB..
head(pae1$plddt) 
```

We can plot the N by N (where N is the number of residues) PAE scores with ggplot or with functions from the Bio3D package:

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

Residue conservation from alignment file:

```{r}
aln_file <- list.files(path=results_dir,
                       pattern=".a3m$",
                        full.names = TRUE,
                       recursive = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```


How many sequences are in this alignment?
```{r}
dim(aln$ali)
```

We can score residue conservation in the alignment with the `conserv()` function.

```{r}
sim <- conserv(aln)
plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```

Note the conserved Active Site residues D25, T26, G27, A28. These positions will stand out if we generate a consensus sequence with a high cutoff value:
```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq

```


For a final visualization of these functionally important sites we can map this conservation score to the Occupancy column of a PDB file for viewing in molecular viewer programs such as Mol*, PyMol, VMD, chimera etc.

```{r}
m1.pdb <- read.pdb(pdb_files[1])
occ <- vec2resno(c(sim[1:99], sim[1:99]), m1.pdb$atom$resno)
write.pdb(m1.pdb, o=occ, file="m1_conserv.pdb")
```

