---
title: "Class05: Data Vis with ggplot2"
author: "Qiman Hu A17912409"
format: pdf
toc: TRUE
---

## Background
There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like the `plot()` function and add on packages like **ggplot2**.

Let's compare how we make a simple figure with these two systems:

We can use the in-built `cars` dataset::

```{r}
head (cars)

```

```{r}
plot(cars)
```
Before I can use **ggplot22** I need to install it on my computer.
To do this we can use the function `install.package("ggpllot2")`

> **N.B.** We never use  `install.package("ggpllot2")` in our quarto doc (we run it once only in our R console) as it would re-install the package every time we render our report. 

Once installed we need to load up the package into our R brain:

```{r}
library(ggplot2)
```

The main function in the **ggplot2** package is called `ggplot()`
```{r}
ggplot(cars)
```

Every ggplot has at least 3 layers:

-the **data** (a data.frame of the stuff we ant to plot)

-the **aes**thetics (how the data maps to the plot)

-the **geom** layer (how you want the plot drawn, e.g. points, lines, etc.)

```{r}
ggplot(cars) +
  aes(x= speed , y= dist) +
  geom_point()
```

## Add some custom features

Let's add a trend line that shows the relationship between speed and distance. 

```{r}
ggplot(cars) +
  aes(x= speed , y= dist) +
  geom_point() +
  geom_smooth()+
  theme_bw()+
  labs(title= "Stopping distance of old cars", x= "Speed (MPH)", y="Distance (ft)")
```

>Q. Can you make the `geom_smooth()` function produce a linear straight line fit to the data and turn off the "gray" error region?

```{r}
ggplot(cars) +
  aes(x= speed , y= dist) +
  geom_point() +
  geom_smooth(methods= "lm", se= FALSE)+
  theme_bw()+
  labs(title= "Stopping distance of old cars", x= "Speed (MPH)", y="Distance (ft)")
```

## Gene Expression

Import the data from course website

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
sum(genes$State == "up")
```

A useful new function in this context is the `table()` function:

```{r}
table(genes$State)
```


My first plot attempt

```{r}
ggplot(genes) +
  aes(x= Condition1, y= Condition2, col=State)+
  geom_point() +
  labs(title = "Gene Expression Changes Upon Drug Treatment", x= "Control (no drug treatment)", y= "Drug Treatment") +
  scale_color_manual(values= c("Red","gray", "Blue"))+
  theme_bw()
```
```{r}
table(genes$State)
```

## Going further

here we read the famous gapminder dataset:

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)

head(gapminder)
```
 > Q.How many entries (rows) are in this dataset?
 
```{r}
nrow(gapminder)
```
 
 > Q. how many "country" entries are there?
 
```{r}
table(gapminder$country)
```
 
```{r}
unique(gapminder$country)
```
```{r}
length(unique(gapminder$country))
```
 
 Lets make our first plot of the entire dataset:
 Plot of "gdpPercap" VS "lifeExp“ colored by "continent"
 
```{r}
p <- ggplot(gapminder)+
aes(gdpPercap, lifeExp, col= continent )+
geom_point(alpha= 0.3)
```
```{r}
p
```
 
```{r}
p+
  facet_wrap(~continent)
```

Make a plot for years 1977 and 2007 only.

> Q. First use the **dplyr** package and the `filter()` function from that package to extract the rows from the year 2007

```{r}
library(dplyr)
g07 <- gapminder %>% filter(year==2007)


```

```{r}
ggplot(g07)+
  aes(gdpPercap, lifeExp, col=continent, size=pop) +
  geom_point( )
```

Do the same for 1977
```{r}
g77 <- gapminder %>% filter(year==1977)
g <- filter(gapminder, year== 2007 | year== 1977)
```

```{r}
ggplot(g)+
  aes(gdpPercap, lifeExp, col=continent, size=pop) +
  geom_point( )+
  facet_wrap(~year)
```


>Q. Make a histogram of life Exp colored by continent (try `fill=continent` or `col=continent`)

```{r}
ggplot(gapminder)+
  aes(lifeExp, col=continent) +
  geom_histogram()
```
```{r}
ggplot(gapminder)+
  aes(lifeExp, fill=continent) +
  geom_histogram()
```

>Q. Make a histogram of life Exp faceted by continent

```{r}
ggplot(gapminder)+
  aes(lifeExp) +
  geom_histogram()+
  facet_wrap(~continent)
```

