---
title: "Class12 HW"
author: "Qiman A17912409"
format: pdf
---


One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.
```{r}
url <- "https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt"

genotype <- read.table(url, header = TRUE)

head(genotype)
```

>Q13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.


```{r}
nrow(genotype)
table(genotype$geno)
```

The sample size for each genotype is as follows: A/A=108, A/G=233. G/G=121

```{r}
library(dplyr)

genotype %>%
  group_by(geno) %>%
  summarize(
    sample_size = n(),
    median_expression = median(exp)
  )
```
The median expression is A/A=31.25, A/G=25.06, G/G=21.07

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(genotype, aes(x = geno, y = exp, fill = geno)) +
  geom_boxplot() +
  labs(
    x = "Genotype",
    y = "ORMDL3 expression",
    title = "ORMDL3 expression by genotype"
  )
```

The boxplot shows that A/A has the highest ORMDL3 expression. This suggests that the SNP is associated with ORMDL3 expression. The G allele appears to lower expression.