---
title: "Class 7 Machine Learning 1"
author: "Qiman A17912409"
format: pdf
toc: true
---

## Background

Today we will explore some core machine learning methods that are very popular in bioinformatics. These include **clustering** and **dimensionallity reduction**. 

## K-means clustering

The main function in "base" R for K-means clustering is called `kmeans()`

Before we go too deep let's make up some "simple" data that we can cluster. To do this we can use the `rnorm()` function:

```{r}
hist(rnorm(1000,mean=3, sd=))
```

```{r}
x <- c(rnorm(30,-3),rnorm(30,+3))
z <- cbind(x=x,y=rev(x))
#rev(x)
plot(z)
```

Now we can run `kmeans()` on this input `z` and see what the results look like.

```{r}
km <- kmeans(z, centers=2)
km
```

```{r}
attributes(km)
```

>Q How many points are in each cluster?

```{r}
km$size
```
>Q. what componnent of your result object details cluster center?

```{r}
km$centers
```
>Q. what componnent of your result object details cluster assignment/membership?

```{r}
km$cluster
```

>Q. PLot `z` colored by the kmeans cluster assignment and add cluster centers as blue points

```{r}
plot(z, col=km$cluster)
points(km$centers, col="blue", pch=15)
```

>Q. Run a K-means clustering and plot the results asking for 4 clusters

```{r}
km4 <- kmeans(z, centers=4)
km4
```

```{r}
plot(z, col=km4$cluster)
points(km4$centers, col="blue", pch=15, cex=2)
```

>**N.B** You need to tell K-means the number of clusters!!

One approach is to try different values for `centers` and then pick the best
```{r}
ans <- NULL
for(i in 1:10){
km <- kmeans(z, centers=i)
ans <- c(ans,km$tot.withinss)
}

ans

plot(ans,typ="o", xlab="Number of clusters", ylab="Total Sum of Squars Distance")
```

## Hierarchical Clustering

The main function in "base" R for Hierarchical Clustering is called `hclust()`

This function does not take your raw data for clustring. You must first build a "distance matrix" from your data and pass this as input to `hclust()`

```{r}
d <- dist(z)
hc <- hclust(d)
hc
```

There is a bespoke`plot()` method for `hclust()` result objects.

```{r}
plot(hc)
abline(h=8,col="red")
```

Once we have our `hclust` object (our tree of cluster dendrogram) we can **cut** the tree to reveal the clustering pattern.

```{r}
cutree(hc, h=8)
```

```{r}
cutree(hc, k=4)
```

>Q make a plot of `z` with your hclust results(i.e. colored by cluster membership)

```{r}
height <- cutree(hc, k=2)
plot(z, col=height)
```

##Principal Component Analysis(PCA)

PCA is a dimensionallity reduction method that is popular for revealing patterns in complex datasets.

### Analysis of UK food data

Let's look at some data on the eating habits of folks from the UK to see if there are patterns and trends that have some regions being distinct from others.

##Data Import

The data is made available in CSV format so we can use the `read.csv()` function for import to R:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

>Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
nrow(x)
ncol(x)
```
There are 17 rows and 5 columns.


## Tidy the data
Fix anything that went wrong with the data import
```{r}
head(x)
```

There seems to be 5 columns instead of 4, we will fix that below:

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```

Now let's check the dimensions again to make sure

```{r}
dim(x)
```

>Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer the second approach where we read the dat file and set the row.names to be the first column. This approach seems better since it avoids problems with running the first approach code block multiple times. We just set the number of rows as 4 from the start.

## Exploratory Analysis
Make some plots to help make sense of obvious trends. We will start with a comparitive bar plot.

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

>Q3: Changing what optional argument in the above barplot() function results in the following plot?

Changing `beside=T` to `beside=F` in the barplot() code results in the following plot. 
```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```
>Q4: Changing what optional argument in the above ggplot() code results in a stacked barplot figure?

Changing the geom_col(position="dodge") to position="stack"


These plots don't really help, we can try a pairs plot.

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

>Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

The pairs function generates a matrix of scatterplotes comparing all pairs of variables in the dataset. Each panel shows the relationship between two countries. A point on the diagonal indicates that the value for that observation is the same in both variables.This indicates identical consumption for that item.

Another useful visualization approach is a **heatmap**

```{r}

heatmap( as.matrix(x) )
```

>Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

According to the figures, England and Whales cluster together, this suggests that their consumption patterns are more similar. I can't easily tell what the mian differences between N.Ireland and the other coountries are.

##PCA to the rescue
The mamin function in "base" R for PCA is called `prcomp()`. This function wants the "obsesrvations" to be in rows and the "variables" to be columns.

So here we need to take the tranpose of our `x` input object:
```{r}
t(x)
```


```{r}
pca <- prcomp(t(x))
summary(pca)
```

To make our new PCA plot we access `pca$x`
```{r}
plot(pca$x[,1], pca$x[,2])
text(pca$x[,1], pca$x[,2], colnames(x))
```

>Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
library(ggplot2)
# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

>Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.


```{r}
# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(aes(label = df$Country),
    vjust = -0.5,
    color = c("red", "blue", "green", "purple")) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

>Q9. Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

Soft drinks and fresh potatoes feature most prominantely. PC2 likely represents variation in countries with higher consuption of soft drinks and potatoes compared to countries with lower consumption of those foods. 

```{r}
# Generate a similar loadings plot for PC2
ggplot(pca$rotation) +
  aes(x = PC2,
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```



